The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC projects 33620180100855CB and 33620180100465CB, and Conicet.}

### 2023-10-04 Independence_CH/axioms_metatheory.tex
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

\subsection{The metatheory Pure}
\begin{isabelle}
Pure.abstract\_rule: (⋀x. ?f(x) ≡ ?g(x)) ⟹ λx. ?f(x) ≡ λx. ?g(x)\isanewline
Pure.combination: ?f ≡ ?g ⟹ ?x ≡ ?y ⟹ ?f(?x) ≡ ?g(?y)\isanewline
Pure.equal\_elim: PROP ?A ≡ PROP ?B ⟹ PROP ?A ⟹ PROP ?B\isanewline
Pure.equal\_intr: (PROP ?A ⟹ PROP ?B) ⟹ (PROP ?B ⟹ PROP ?A) ⟹ \isanewline
\ \ \ \ \ \ \ \ \ \ PROP ?A ≡ PROP ?B\isanewline
Pure.reflexive: ?x ≡ ?x\isanewline
Pure.symmetric: ?x ≡ ?y ⟹ ?y ≡ ?x\isanewline
Pure.transitive: ?x ≡ ?y ⟹ ?y ≡ ?z ⟹ ?x ≡ ?z
\end{isabelle}

\subsection{IFOL and FOL}
In the axioms \isa{refl}, \isa{subst}, \isa{allI}, \isa{spec},
\isa{exE}, \isa{exI}, and \isa{eq\_reflection} there is a constraint
for the \emph{type} of the variables \isa{a, b, x} to be in the class \isa{term\_class}.
\begin{isabelle}
IFOL.FalseE: ⋀P. False ⟹ P\isanewline
IFOL.refl:  (⋀a. a = a)\isanewline
IFOL.subst: (⋀a b P. a = b ⟹ P(a) ⟹ P(b))\isanewline
IFOL.allI:  (⋀P. (⋀x. P(x)) ⟹ ∀x. P(x))\isanewline
IFOL.spec:  (⋀P x. ∀x. P(x) ⟹ P(x))\isanewline
IFOL.exE:   (⋀P R. ∃x. P(x) ⟹ (⋀x. P(x) ⟹ R) ⟹ R)\isanewline
IFOL.exI:   (⋀P x. P(x) ⟹ ∃x. P(x))\isanewline
IFOL.conjI: ⋀P Q. P ⟹ Q ⟹ P ∧ Q\isanewline
IFOL.conjunct1: ⋀P Q. P ∧ Q ⟹ P\isanewline
IFOL.conjunct2: ⋀P Q. P ∧ Q ⟹ Q\isanewline
IFOL.disjE: ⋀P Q R. P ∨ Q ⟹ (P ⟹ R) ⟹ (Q ⟹ R) ⟹ R\isanewline
IFOL.disjI1: ⋀P Q. P ⟹ P ∨ Q\isanewline
IFOL.disjI2: ⋀P Q. Q ⟹ P ∨ Q\isanewline
IFOL.eq\_reflection: (⋀x y. x = y ⟹ x ≡ y)\isanewline
IFOL.iff\_reflection: ⋀P Q. P ⟷ Q ⟹ P ≡ Q\isanewline
IFOL.impI: ⋀P Q. (P ⟹ Q) ⟹ P ⟶ Q\isanewline
IFOL.mp: ⋀P Q. P ⟶ Q ⟹ P ⟹ Q\isanewline%\isanewline
FOL.classical: ⋀P. (¬ P ⟹ P) ⟹ P
\end{isabelle}

>>>>

\subsection{The metatheory $\Meta$}
Below, the \isatt{PROP} operator corresponds to the
injection $[\cdot]$ of Section~\ref{sec:isabelle-metalogic-meta}.
\begin{isabelle}
Pure.abstract\_rule: (⋀x. ?f(x) ≡ ?g(x)) ⟹ λx. ?f(x) ≡ λx. ?g(x)\isanewline
Pure.combination: ?f ≡ ?g ⟹ ?x ≡ ?y ⟹ ?f(?x) ≡ ?g(?y)\isanewline
Pure.equal\_elim: PROP ?A ≡ PROP ?B ⟹ PROP ?A ⟹ PROP ?B\isanewline
Pure.equal\_intr: (PROP ?A ⟹ PROP ?B) ⟹ (PROP ?B ⟹ PROP ?A) ⟹ \isanewline
\ \ \ \ \ \ \ \ \ \ PROP ?A ≡ PROP ?B\isanewline
Pure.reflexive: ?x ≡ ?x\isanewline
Pure.symmetric: ?x ≡ ?y ⟹ ?y ≡ ?x\isanewline
Pure.transitive: ?x ≡ ?y ⟹ ?y ≡ ?z ⟹ ?x ≡ ?z
\end{isabelle}

\subsection{IFOL and FOL}
In the axioms \isa{refl}, \isa{subst}, \isa{allI}, \isa{spec},
\isa{exE}, \isa{exI}, and \isa{eq\_reflection} there is a constraint
on the \emph{type} of the variables \isa{a, b, x}: It should be in the class \isa{term\_class}.
\begin{isabelle}
IFOL.FalseE: ⋀P. False ⟹ P\isanewline
IFOL.refl:  (⋀a. a = a)\isanewline
IFOL.subst: (⋀a b P. a = b ⟹ P(a) ⟹ P(b))\isanewline
IFOL.allI:  (⋀P. (⋀x. P(x)) ⟹ ∀x. P(x))\isanewline
IFOL.spec:  (⋀P x. ∀x. P(x) ⟹ P(x))\isanewline
IFOL.exE:   (⋀P R. ∃x. P(x) ⟹ (⋀x. P(x) ⟹ R) ⟹ R)\isanewline
IFOL.exI:   (⋀P x. P(x) ⟹ ∃x. P(x))\isanewline
IFOL.conjI: ⋀P Q. P ⟹ Q ⟹ P ∧ Q\isanewline
IFOL.conjunct1: ⋀P Q. P ∧ Q ⟹ P\isanewline
IFOL.conjunct2: ⋀P Q. P ∧ Q ⟹ Q\isanewline
IFOL.disjE: ⋀P Q R. P ∨ Q ⟹ (P ⟹ R) ⟹ (Q ⟹ R) ⟹ R\isanewline
IFOL.disjI1: ⋀P Q. P ⟹ P ∨ Q\isanewline
IFOL.disjI2: ⋀P Q. Q ⟹ P ∨ Q\isanewline
IFOL.eq\_reflection: (⋀x y. x = y ⟹ x ≡ y)\isanewline
IFOL.iff\_reflection: ⋀P Q. P ⟷ Q ⟹ P ≡ Q\isanewline
IFOL.impI: ⋀P Q. (P ⟹ Q) ⟹ P ⟶ Q\isanewline
IFOL.mp: ⋀P Q. P ⟶ Q ⟹ P ⟹ Q\isanewline%\isanewline
FOL.classical: ⋀P. (¬ P ⟹ P) ⟹ P
\end{isabelle}

### 2023-10-04 Independence_CH/cofinality.tex
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

It turns out that the rest of the basic results on cofinality (namely,
idempotence of $\cf$, that regular ordinals are cardinals, the
cofinality of Alephs, König's Theorem) follow easily from the previous
Lemma by “algebraic” reasoning only.

We therefore expect that the relativization of these
results be straightforward, when time permits.

>>>>

It turns out that the rest of the basic results on cofinality (namely,
idempotence of $\cf$, that regular ordinals are cardinals, the
cofinality of Alephs, König's Theorem) follow easily from the previous
Lemma by “algebraic” reasoning only.
We expect the relativization of these
results to be straightforward.

The AFP entry \cite{Delta_System_Lemma-AFP} also includes the
formalization of the (absolute) Delta System Lemma (DSL). Formalizing its
proof was rather straightforward, once the many prerequisites were
taken care of. Some of those were really basic, for instance:
\begin{enumerate}
\item \label{item:1}$\omega$ injects into every infinite set;
\item \label{item:2}surjective images of countable sets are countable;
\item \label{item:3}the union over a countable index set $J$ of a family $X :: \tyi
  \fun \tyi$ of countable sets is countable.
\end{enumerate}
It was also convenient to isolate the relevant recursive construction
principle (\isatt{bounded{\uscore}cardinal{\uscore}selection}) that
appears in the proof of DSL, which was also useful for showing
Item~\ref{item:1}:
\begin{lemma}\label{lem:bdd-card-selection}
  Assume $F$ is nonempty,
  $\gamma$ is a cardinal, and $Q$ is a binary predicate over $F$ satisfying 
  \[
    \forall Y \sbq F.\ |Y| < \gamma \implies \exists a\in F.\ \forall
    y\in Y.\ Q(y,a).
  \]
  Then there exists  $S:\gamma \to F$ such that
  $Q(S(\alpha),S(\beta))$ for all $\alpha<\beta<\gamma$.
\end{lemma}

Concerning the relativization of the proof of DSL for its use in
forcing, it required inserting proofs that all the relevant objects
lied in the model $M$; this was only tedious.
A bit more effort was required at the point where Item~\ref{item:3}
was used, because it involved an application of Replacement (being $X$
a class function); relativizing
Lemma~\ref{lem:bdd-card-selection} also required some work, because of the
recursion.

### 2023-10-04 Independence_CH/lessons.tex
Explanation of separation, strong_repl, blowup, ack Commelin
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

Nearly the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example,
\begin{itemize}
\item
  we only proved the Delta System Lemma for $\aleph_1$-sized families;
  thus limiting us to the case of the $\aleph_1$-chain condition, and avoiding the relativization
  of the material on cofinalities \cite{Delta_System_Lemma-AFP};
\item we showed preservation of sequences by considering countably
  closed forcings (in fact, we formalized the bare minimum requirement
  of being $({<}\omega{+}1)$-closed, that is, closure under
  $\omega$-sequences and not $\delta$-sequences for every countable $\delta$).
\end{itemize}
In doing this we went against the conventional wisdom that one should formalize the most general version of the
results available. Another shortcut we took was to
simplify some proofs by appealing to the countability of the ground
model; this is the case of
\isa{definition{\uscore}of{\uscore}forces} and the result on forcing
values of a function.

@@

Nevertheless, experience  in software engineering is invaluable in
large projects like the present one. For example, it is
(mathematically) misleading when automatic tools (\isa{simp},
\isa{auto}, etc) stop working just because of the sheer size of the
goal (v.g., the same statement with 7 variables succeeds but with 8
variables does not). Scale issues are very easily disregarded in the
abstract but, as a colorful example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
can be explicitly printed by Isabelle2021-1 (it spans nearly 20k symbols), but
$\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$
can not.

>>>>

Upon reaching the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example,
\begin{itemize}
\item
  we only proved the Delta System Lemma for $\aleph_1$-sized families;
  thus limiting us to the case of the $\aleph_1$-chain condition, and avoiding the relativization
  of the material on cofinalities \cite{Delta_System_Lemma-AFP};
\item we showed preservation of sequences by considering countably
  closed forcings (in fact, we formalized the bare minimum requirement
  of being $({<}\omega{+}1)$-closed, that is, closure under
  $\omega$-sequences and not $\delta$-sequences for every countable $\delta$).
\end{itemize}
In doing this we went against the conventional wisdom that one should formalize the most general version of the
results available. Another shortcut we took was to
simplify some proofs by appealing to the countability of the ground
model; this is the case of
\isa{definition{\uscore}of{\uscore}forces} and the result on forcing
values of a function.

@@

Nevertheless, experience  in software engineering is invaluable in
large projects like the present one. For example, it is
(mathematically) misleading when automatic tools (\isa{simp},
\isa{auto}, etc) stop working just because of the sheer size of the
goal (v.g., the same statement with 7 variables succeeds but with 8
variables does not). Scale issues are very easily disregarded in the
abstract but, as a colorful example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
can be explicitly printed by Isabelle2021-1 (it spans nearly 20k symbols), but
$\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$
can not.
(For this particular example, one reason for the blow-up in
length comes from the encoding of negation using the “nand”
connective, but otherwise the formulas grow bigger by the sheer
complexity of the definition of the forcing relation.)

### 2023-10-04 Independence_CH/proof_assists_isabelle.tex
Explanation of Hessenberg and minor edits
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

Isabelle/ZF reaches essentially Hessenberg's $|A|\cdot|A| = |A|$. Our decision
(during 2017) to
use this assistant was triggered by its constructibility
library, \session{ZF-Constructible} \citep{paulson_2003},
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/
which contains the development of $L$, the proof that it satisfies
$\AC$, and a version of the Reflection Principle. The latter was
actually encoded as a series of instructions to Isabelle automatic
proof tools that would prove each particular instance of reflection:
This is an example of what was said at the end of Section~\ref{sec:isabelle-metalogic-meta}.

>>>>

The amount of set theory developed in “vanilla” Isabelle/ZF is quite
modest, essentially going no further than Hessenberg's theorem on the
cardinality of products $|A\times A|=|A|\cdot|A| = |A|$. Instead, our decision
(during 2017) to
use this assistant was triggered by its constructibility
library, \session{ZF-Constructible} \citep{paulson_2003},
% https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/
which contains the development of $L$, the proof that it satisfies
$\AC$, and a version of the Reflection Principle. The latter was
actually encoded as a series of instructions to Isabelle automatic
proof tools that would prove each particular instance of reflection:
This is an example of what was said at the end of Section~\ref{sec:isabelle-metalogic-meta}.

### 2023-10-04 Independence_CH/related.tex
Probably final revision
Co-authored-by: Miguel Pagano <miguel.pagano@gmail.com>

This is perhaps an appropriate time to insist that we have \emph{not}
formalized the relative consistency of $\neg\CH$, and we are actually
not aiming for that in the short term. In our context, going for the
plain consistency result (that is, proving the arithmetical statement
$\Con(ZFC) \implies \Con(ZFC + \neg\CH)$ over a weak base theory)
seems off the mark, since our base is fixed (and essentially equivalent to $\ZF$). %%,
%% and we catch our tail.
If we intended to do so, we would a need a development of a weak base with a
new type of (external) formulas having the required
induction principles, plus an interface from this new type to
Isabelle/ZF. This might allow us to take the standard route to use ctms for a
relative consistency proof (through a version of the Reflection
Theorem proved by induction on the external formulas).

>>>>

This is perhaps an appropriate time to insist that we have \emph{not}
formalized the relative consistency of $\neg\CH$, and we are actually
not aiming for that in the short term. In our context, going for the
plain consistency result (that is, proving the arithmetical statement
$\Con(ZFC) \implies \Con(ZFC + \neg\CH)$ over a weak base theory)
seems off the mark, since our base is fixed (and essentially equivalent to $\ZF$). %%,
%% and we catch our tail.
If we intended to do so, we would need a development of a weak base with a
new type of (external) formulas having the required
induction principles, plus an interface from this new type to
Isabelle/ZF. This might allow us to take the standard route to use ctms for a
relative consistency proof (through a version of the Reflection
Theorem proved by induction on the external formulas).
